---
title: "ggden — DEN House-Style Charts for ggplot2"
subtitle: "Dewan Ekonomi Nasional"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
  fig-dpi: 300
---

## Installation

Install directly from GitHub:

```r
# install.packages("remotes")  # if needed
remotes::install_github("den-econ/ggden")
```

## Setup

```{r}
library(ggden)
library(ggplot2)
library(dplyr)
library(tidyr)
```

## 1. Palette Preview

The DEN palette contains 9 colours. Access them with `den_palette()` or
individually via `den_colors["gold"]`.

```{r}
#| fig-width: 8
#| fig-height: 1.5
pal <- den_palette()
df_pal <- data.frame(
  color = factor(names(pal), levels = names(pal)),
  hex   = unname(pal),
  y     = 1
)

ggplot(df_pal, aes(x = color, y = y, fill = color)) +
  geom_tile(width = 0.9, height = 0.9) +
  geom_text(aes(label = hex), vjust = 2.5, size = 3) +
  scale_fill_manual(values = pal) +
  theme_void() +
  theme(
    axis.text.x = element_text(size = 9, face = "bold", margin = margin(t = 4)),
    legend.position = "none"
  ) +
  labs(title = "DEN Palette") +
  coord_fixed(ratio = 0.6)
```

## 2. Line Chart — GDP Growth

A multi-series line chart with `theme_den()` and `scale_color_den()`.
Equivalent to the GDP growth chart in `fig.ipynb`.

```{r}
#| fig-width: 9
#| fig-height: 6
set.seed(42)
years <- 2015:2024

df_gdp <- data.frame(
  Tahun = rep(years, 4),
  Industri = rep(
    c("KPK dan Alas Kaki", "TPT",
      "Industri Manufaktur Non Migas", "Nasional"),
    each = length(years)
  ),
  Pertumbuhan = c(
    cumsum(rnorm(length(years), 3, 4)) / 5,
    cumsum(rnorm(length(years), 2, 3)) / 5,
    cumsum(rnorm(length(years), 4, 2)) / 5,
    cumsum(rnorm(length(years), 5, 1)) / 5
  )
)

ggplot(df_gdp, aes(x = Tahun, y = Pertumbuhan, colour = Industri)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60", alpha = 0.5) +
  scale_x_continuous(breaks = years) +
  scale_color_den() +
  theme_den() +
  labs(y = "Pertumbuhan (%)", x = NULL)
```

## 3. Dual-Panel Line Chart — COVID Disruption

Two side-by-side panels showing actual GDP vs. a pre-COVID trend.

```{r}
#| fig-width: 12
#| fig-height: 4.5
set.seed(7)
years2 <- 2010:2024

df_dual <- data.frame(
  Tahun = rep(years2, 2),
  Sektor = rep(c("TPT", "KPK dan Alas Kaki"), each = length(years2)),
  GDP = c(
    cumsum(rnorm(length(years2), 500, 100)) + 5000,
    cumsum(rnorm(length(years2), 300, 80))  + 3000
  )
)
# simulate COVID dip
df_dual$GDP[df_dual$Tahun == 2020] <- df_dual$GDP[df_dual$Tahun == 2020] * 0.85

# pre-COVID trend (simple linear fit on 2010-2019)
df_trend <- df_dual |>
  filter(Tahun <= 2019) |>
  group_by(Sektor) |>
  summarise(
    intercept = coef(lm(GDP ~ Tahun))[1],
    slope     = coef(lm(GDP ~ Tahun))[2],
    .groups   = "drop"
  )

df_trend_line <- expand.grid(Tahun = years2, Sektor = unique(df_dual$Sektor)) |>
  left_join(df_trend, by = "Sektor") |>
  mutate(Trend = intercept + slope * Tahun)

ggplot() +
  geom_line(data = df_dual, aes(x = Tahun, y = GDP, colour = "PDB Aktual"),
            linewidth = 1.2) +
  geom_point(data = df_dual, aes(x = Tahun, y = GDP, colour = "PDB Aktual"),
             size = 1.5) +
  geom_line(data = df_trend_line,
            aes(x = Tahun, y = Trend, colour = "Tren Pra-COVID"),
            linewidth = 1, linetype = "dashed", alpha = 0.8) +
  geom_vline(xintercept = 2020, colour = "red", alpha = 0.5) +
  facet_wrap(~Sektor, scales = "free_y") +
  scale_x_continuous(breaks = seq(2010, 2024, 2)) +
  scale_colour_manual(values = c(
    "PDB Aktual"     = den_colors["gold"],
    "Tren Pra-COVID" = den_colors["brown"]
  )) +
  theme_den() +
  labs(y = "GDP (Milyar Rupiah)", x = NULL)
```

## 4. Stacked Bar Chart — Export by Region

Stacked bars with percentage annotations, matching the export-by-region chart.

```{r}
#| fig-width: 10
#| fig-height: 6
df_export <- data.frame(
  Year   = rep(2022:2024, each = 4),
  Region = rep(c("USA", "EU28", "Asia", "RoW"), 3),
  Value  = c(2.1, 1.8, 3.0, 1.0,
             2.3, 1.9, 3.2, 1.1,
             2.5, 2.0, 3.5, 1.2)
)

# compute share labels
df_export <- df_export |>
  group_by(Year) |>
  mutate(
    Share = Value / sum(Value),
    Label = sprintf("%.1f%%", Share * 100),
    ypos  = cumsum(Value) - Value / 2
  ) |>
  ungroup()

ggplot(df_export, aes(x = factor(Year), y = Value, fill = Region)) +
  geom_col(width = 0.7) +
  geom_text(aes(y = ypos, label = Label),
            colour = "white", fontface = "bold", size = 3.5) +
  scale_fill_den() +
  theme_den() +
  labs(y = "Nilai ekspor (Milyar USD)", x = NULL)
```

## 5. Simple Bar Chart — Trade Values (Grid)

A 3x2 faceted grid of bar charts, one per trade type and flow direction.

```{r}
#| fig-width: 10
#| fig-height: 10
set.seed(99)
types  <- c("Bahan Baku", "Barang Antara", "Barang Jadi")
flows  <- c("Impor", "Ekspor")

df_trade <- expand.grid(
  Tahun = 2022:2024,
  Tipe  = types,
  Flow  = flows
) |>
  mutate(Value = runif(n(), 1, 8))

# assign label per facet
facet_labels <- paste0(
  "(", letters[1:6], ") ",
  rep(flows, times = 3), " ",
  rep(types, each = 2)
)
df_trade <- df_trade |>
  mutate(
    Facet = paste0(Flow, " ", Tipe),
    Facet = factor(Facet, levels = paste0(
      rep(flows, times = 3), " ", rep(types, each = 2)
    ))
  )

ggplot(df_trade, aes(x = factor(Tahun), y = Value,
                     fill = Flow)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.2f", Value)),
            vjust = -0.3, fontface = "bold", size = 3) +
  facet_wrap(~Facet, ncol = 2, scales = "fixed") +
  scale_fill_manual(values = c(
    "Impor"  = den_colors["gold"],
    "Ekspor" = den_colors["brown"]
  )) +
  theme_den() +
  labs(y = "Nilai (Miliar USD)", x = NULL)
```

## 6. Grouped Bar Chart — Import Shares

Side-by-side bars comparing two years.

```{r}
#| fig-width: 10
#| fig-height: 6
df_share <- data.frame(
  Partner = rep(c("CHN", "VNM", "IND", "BGD", "IDN", "MEX"), each = 2),
  Year    = rep(c("2015", "2024"), 6),
  Share   = c(44.0, 29.2, 11.8, 19.2, 5.1, 6.5,
              4.4, 6.0, 5.0, 5.5, 3.8, 3.7)
)
df_share$Partner <- factor(df_share$Partner,
                           levels = c("CHN", "VNM", "IND", "BGD", "IDN", "MEX"))

ggplot(df_share, aes(x = Partner, y = Share, fill = Year)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(
    aes(label = sprintf("%.1f%%", Share)),
    position = position_dodge(width = 0.7),
    vjust = -0.4, fontface = "bold", size = 3
  ) +
  scale_fill_den() +
  theme_den() +
  labs(
    y     = "Import Share (%)",
    x     = NULL,
    title = "USA Textile Import Share by Partner"
  )
```

## 7. Using the Theme with Any ggplot

`theme_den()` works as a drop-in replacement for any ggplot2 theme.
Just add it to your plot:

```{r}
#| fig-width: 8
#| fig-height: 5
ggplot(mpg, aes(x = displ, y = hwy, colour = factor(cyl))) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_den() +
  theme_den() +
  labs(
    title    = "Engine Displacement vs Highway MPG",
    subtitle = "By number of cylinders",
    x = "Displacement (L)",
    y = "Highway MPG",
    colour = NULL
  )
```

## Quick Reference

| Function | Purpose |
|---|---|
| `theme_den()` | Apply DEN theme to any ggplot |
| `scale_color_den()` | DEN discrete colour scale |
| `scale_fill_den()` | DEN discrete fill scale |
| `den_palette(n)` | Get first `n` DEN colours |
| `den_colors["gold"]` | Access a single named colour |
| `den_save("file.png", p)` | Save at 300 dpi with white background |
